---
title: "outbreakdistr"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This package contains functions for computing the probability distributions for the sizes of
outbreaks of communicable diseases in closed groups. The methods are exact, meaning they don't rely on simulation methods. 
The package is mostly suitable for smaller outbreaks (households, schools, etc), but not outbreaks in larger groups
such as cities or countries. This is because the methods tend to be slow and numerically unstable for larger groups.

This package is in early stages of development. Functionality, interface and algorithmic details will probably change substantially. 

Currently implemented:

- Final Size methods from Ball (1986): Single type and multitype populations, with different infectious period distributions.
- Epidemic Markov Models: Continuous time SIR and SIS models, discrete time SIR. 


Keywords: Infectious disease modelling, Epidemic modelling, Final size distribution.

## Installation

To install the latest development version:

```{r eval=FALSE}
devtools::install_github("opisthokonta/outbreakdistr")
```


## Examples

First we must load the package:
```{r}
library(outbreakdistr)
```

### Final Size distribution with general infectious period

The method from Ball (1986) is implemented in the fsdistr function. Here is how to 
compute the final size distribution with infection rate beta = 1.5, and the infectious
period is exponentially distributed with rate  gamma = 1, in a population of 7 people, 
with two primary cases and 5 initially susceptible. This corresponds to a
basic reproduction number R0  = 1.5.

```{r}

fs_res1 <- fsdistr(s0 = 5, i0 = 2, beta = 1.6, ip_model = 'exponential', ip_params = c(1.0))

barplot(height = fs_res1, names.arg = 0:5, xlab = 'Number of secondary cases', ylab = 'Probability')

```

Here is the probability distribution if the infectious period is constant, with 
length of 1 unit of time:

```{r}
fsdistr(s0 = 5, i0 = 2, beta = 1.6, ip_model = 'constant', ip_params = c(1.0))
```
and here is the probability distribution if the infectious period follows a gamma distribution, with shape parameter 2 and
rate parameter 3:

```{r}
fsdistr(s0 = 5, i0 = 2, beta = 1.6, ip_model = 'gamma', ip_params = c(2, 3))
```

### Multitype Final Size distribution with general infectious period

An extension to a multitype (multivariate) scenario has also been implemented. In this example
we consider two groups, where the second group has much lower susceptibility (perhaps due to vaccination).
The second group has exponentially distributed infectious period, while the first group has gamma distributed
infectious period. The infectious period parameters need to be specified using a list. The result is a named 
vector showing the joint probability distribution of the number of secondary cases in each group.

```{r}
i0 <- c(1,0) # primary case in first group.
s0 <- c(3,3)

beta_matrix <- matrix(c(1.5,1.5,0.1,0.1), ncol = 2, nrow= 2)

fs_mt_res <- fsdistr_mt(s0 = s0, i0 = i0, beta = beta_matrix, 
                        ip_model = c('gamma', 'exponential'), 
                        ip_params = list(c(3, 2), c(1.1)))

fs_mt_res

```
It can be useful to also look at the marginal distributions of the number of secondary cases in each group. Notice how the group with low susceptibility has much higher chance of no cases.

```{r}
# Group 1
sapply(split(fs_mt_res, f = substr(names(fs_mt_res), start = 1, stop = 1)), FUN = sum)
# Group 2 - low susceptibility
sapply(split(fs_mt_res, f = substr(names(fs_mt_res), start = 3, stop = 3)), FUN = sum)
```


# Markov models

The Epidemic Markov model is a discrete state model where the states are the number of susceptible 
and infectious individuals in the population. The continuous time version of the model is implemented
in the emmct function. Lets take a look at the cumulative incidence a short period after (0.7 infectious periods) the outbreak starts:

```{r}
emmct_res <- emmct(s0 = 5, i0 = 2, beta = 1.6, gamma = 1, time = 0.7)
emmct_res$fs_distr
```
Lets take a look at how it looks some time later (4 infectious periods). Notice how the distribution is 
similar to the one computed in the very first example.

```{r}
emmct_res2 <- emmct(s0 = 5, i0 = 2, beta = 1.6, gamma = 1, time = 4)
emmct_res2$fs_distr

# compare withe final size distribution with exponential infectious period
fs_res1
```

The discrete time version of the Epidemic Markov Model (function emmdt) can be used to compute the final size distribution, which should match the result from fsdistr. The emmdt is more numerically stable for larger population sizes than fsdistr in my experience. 


```{r}
emmdt_res <- emmdt(s0 = 5, i0 = 2, beta = 1.6, gamma = 1)
emmdt_res$fs_distr
```

## Precision arithmetic

The exact methods can sometimes be numerically unstable, yielding negative probabilities. It is not easy to predict exactly which situations will give instability, but tend to occur for population sizes greater than 20, and when a large portion of the sample space have probabilities close to 0. It can be remedied by using higher precision arithemtic. Here is an example of a setting giving negative probabilties. A warning is given, suggesting using higher precision arithmetic.



```{r}

# fs_res_unstable <- fsdistr(s0 = 50, i0 = 1, beta = 1.4, ip_model = 'exponential', ip_params = c(1.0))
fs_res_unstable <- fsdistr(s0 = 50, i0 = 1, beta = 1.2, ip_model = 'exponential', ip_params = c(1.0))
barplot(height = fs_res_unstable, names.arg = 0:50, xlab = 'Number of secondary cases', ylab = 'Probability')
```

Use the 'prec' argument to set the precision. Try setting it to 64, and increase it if it doesn't work. The precision arithmetic is provided trough the Rmpfr package, and the prec argument corresponds to the precBits argument used in Rmpfr.
```{r}
fs_res_precise <- fsdistr(s0 = 50, i0 = 1, beta = 1.2, ip_model = 'exponential', ip_params = c(1.0), prec = 64)
barplot(height = fs_res_precise, names.arg = 0:50, xlab = 'Number of secondary cases', ylab = 'Probability')
```

Numerical issues can also happen without the probabilities becoming negative. Here is an example for a 45 person population, where we zoom in on the probability distribution from 20 to 45 cases. We see that it's a bit jaggged, which is unrealistic. Increasing the arithmetic precision gives a smoother and more realistic probability distribution. 

Since none of the probabilities are negative, it does not throw a warning, and I don't know how to identify these signs of numerical instability, other than visual inspection. 

```{r}
fs_res_unstable2 <- fsdistr(s0 = 45, i0 = 1, beta = 1.11, ip_model = 'exponential', ip_params = c(1.0))
fs_res_precise2 <- fsdistr(s0 = 45, i0 = 1, beta = 1.11, ip_model = 'exponential', ip_params = c(1.0), prec=64)

layout(matrix(c(1,2), nrow=1))
barplot(fs_res_unstable2[21:46], names.arg = 20:45, main = 'normal precision')
barplot(fs_res_precise2[21:46], names.arg = 20:45, main = 'increased precision')



```


```{r}

```


## Other packages

The [chainbinomial](https://github.com/opisthokonta/chainbinomial) package implements outbreak distributions using Chain Binomial models, including methods for estimation and regression modelling. The [finalsize](https://github.com/epiverse-trace/finalsize) can be used to compute the expected values for models, including larger populations. 



## References

- Ball, Frank (1986) A unified approach to the distribution of total size and total area under the trajectory of infectives in epidemic models







